---
title: "Text Analysis and Spatial Data for Economists"
subtitle: "Urban Green Space and Urban Heat: Multi-City Spatial Analysis"
author: LEONARDO GRAVELLONE
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
  highlight: tango
  df_print: kable
  toc: true
  toc_float: true
  number_sections: true
toc_collapsed: true
 
---


```{r setup, include=F , message = F}
library(dplyr)
library(terra)
library(exactextractr)
library(sf)
library(raster)
library(ggplot2) 
library(tigris)
library(tidycensus)
options(tigris_use_cache = TRUE, tigris_class = "sf")
options(readr.show_progress = FALSE)
options(dplyr.show_progress = FALSE)
options(terra.progress = 0)
```

# *1. Downloading and loading all the necessary data*
```{r load_data, message=FALSE, warning=FALSE}
cities <- tibble::tibble(
  city = c("New York", "Los Angeles", "Chicago", "Houston", "Phoenix",
           "Philadelphia", "San Antonio", "San Diego", "Dallas", "San Jose"),
  state = c("NY", "CA", "IL", "TX", "AZ", "PA", "TX", "CA", "TX", "CA"),
  county_fips = c("061", "037", "031", "201", "013", "101", "029", "073", "113", "085")
)
tract_list <- list()
for (i in seq_len(nrow(cities))) {
  message("Downloading: ", cities$city[i])
  
  tracts_i <- tracts(state = cities$state[i],
                     county = cities$county_fips[i],
                     year = 2020,
                     cb = TRUE) |>
    st_transform(4326) |>           
    mutate(city = cities$city[i])     
  
  tract_list[[cities$city[i]]] <- tracts_i
}
all_tracts <- do.call(rbind, tract_list)
```

# *2. Preprocessing data*

```{r preproc_data, message=FALSE, warning=FALSE, results='hide'}
tracts <- st_read("data/us_10city_tracts_2020.geojson")
cities <- unique(tracts$city)
dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)
tracts$city <- gsub(" ", "", tracts$city)


for (city in unique(tracts$city)) {
  message("Processing: ", city)
  
  city_tracts <- tracts %>% filter(city == !!city)
  ndvi_path <- paste0("data/gee_exports/", city, "_NDVI.tif")
  lst_path  <- paste0("data/gee_exports/", city, "_LST.tif")
  
  if (!file.exists(ndvi_path) | !file.exists(lst_path)) {
    warning("Skipping ", city, ": NDVI or LST file missing")
    next
  }
  ndvi <- rast(ndvi_path)
  lst  <- rast(lst_path)
  ndvi <- project(ndvi, crs(city_tracts))
  lst  <- project(lst, crs(city_tracts))
  ndvi_crop <- crop(ndvi, city_tracts)
  lst_crop  <- crop(lst, city_tracts)
  ndvi_mask <- mask(ndvi_crop, vect(city_tracts))
  lst_mask  <- mask(lst_crop, vect(city_tracts))
  writeRaster(ndvi_mask,
              filename = paste0("data/processed/", city, "_NDVI_masked.tif"),
              overwrite = TRUE)
  
  writeRaster(lst_mask,
              filename = paste0("data/processed/", city, "_LST_masked.tif"),
              overwrite = TRUE)
  
  message("Done: ", city)
}
```

# *3. Extract NDVI & LST per Census Tract*

```{r extract_data, message=FALSE, warning=FALSE, results='hide'}
tracts <- st_read("data/us_10city_tracts_2020.geojson")
tracts$city <- gsub(" ", "", tracts$city)
cities <- unique(tracts$city)
results <- list()

for (city in cities) {
  city_tracts <- tracts %>% filter(city == !!city)
  ndvi_file <- paste0("data/processed/", city, "_NDVI_masked.tif")
  lst_file  <- paste0("data/processed/", city, "_LST_masked.tif")
  
  if (!file.exists(ndvi_file) | !file.exists(lst_file)) {
    next
  }
  ndvi <- rast(ndvi_file)
  lst  <- rast(lst_file)
  city_tracts$mean_ndvi <- exact_extract(ndvi, city_tracts, 'mean')
  city_tracts$mean_lst  <- exact_extract(lst,  city_tracts, 'mean')
  results[[city]] <- city_tracts
}

zonal_stats <- do.call(rbind, results)

st_write(zonal_stats, "data/ndvi_lst_by_tract.geojson", delete_dsn = TRUE)
write.csv(st_drop_geometry(zonal_stats), "data/ndvi_lst_by_tract.csv", row.names = FALSE)
```
## *4. Correlation Table*
```{r sum_table, message=FALSE, warning=FALSE}
zonal <- read.csv("data/ndvi_lst_by_tract.csv")
results <- list()
zonal$city <- gsub(" ", "", zonal$city)

for (city in unique(zonal$city)) {
  df <- zonal %>% filter(city == !!city) %>%
    filter(!is.na(mean_ndvi), !is.na(mean_lst))
  if (nrow(df) < 50) next
  cor_val <- cor(df$mean_ndvi, df$mean_lst)
  model <- lm(mean_lst ~ mean_ndvi, data = df)
  slope <- coef(model)[["mean_ndvi"]]
  r2 <- summary(model)$r.squared
  results[[city]] <- data.frame(
    city = city,
    correlation = round(cor_val, 3),
    slope = round(slope, 2),
    r_squared = round(r2, 3),
    n_tracts = nrow(df)
  )
}
summary_table <- do.call(rbind, results)
summary_table <- summary_table[order(-summary_table$r_squared), ]

knitr::kable(summary_table, digits = 3,
             col.names = c("City", "Correlation", "Slope (°C/NDVI)", "R²", "Number of Tracts"),
             caption = "Summary of NDVI–LST Relationship by City")
write.csv(summary_table, "data/ndvi_lst_summary_by_city.csv", row.names = FALSE)
```
## *5. Correlation Plot by City*
```{r cor_plot, message=FALSE, warning=FALSE}

zonal <- read.csv("data/ndvi_lst_by_tract.csv")
zonal$city <- gsub(" ", "", zonal$city)

zonal_filtered <- zonal %>%
  filter(!is.na(mean_ndvi), !is.na(mean_lst))

ggplot(zonal_filtered, aes(x = mean_ndvi, y = mean_lst)) +
  geom_point(alpha = 0.5, size = 0.8, color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  facet_wrap(~ city, scales = "free_y") +
  labs(title = "NDVI vs. LST by City",
       x = "Mean NDVI",
       y = "Mean LST (°C)") +
  theme_minimal(base_size = 10)
```


## *6. Cooling effect and R^2 Plot*
```{r plot, message=FALSE, warning=FALSE}
summary_table <- read.csv("data/ndvi_lst_summary_by_city.csv")

ggplot(summary_table, aes(x = reorder(city, slope), y = slope)) +
  geom_col(fill = "#3a8d63") +
  coord_flip() +
  labs(title = "Cooling Effect of NDVI (Slope)", x = "City", y = "Slope (°C per +1 NDVI)") +
  theme_minimal()

ggplot(summary_table, aes(x = reorder(city, r_squared), y = r_squared)) +
  geom_col(fill = "#4682b4") +
  coord_flip() +
  labs(title = "R²: Variation in LST Explained by NDVI", x = "City", y = "R²") +
  theme_minimal()
```


## *7. Top and Bottom Decile Difference Table*
``` {r dec_table, message=FALSE, warning=FALSE}
zonal <- read.csv("data/ndvi_lst_by_tract.csv")
zonal$city <- gsub(" ", "", zonal$city)

quartile_summary <- zonal %>%
  filter(!is.na(mean_ndvi), !is.na(mean_lst)) %>%
  group_by(city) %>%
  mutate(ndvi_qtile = ntile(mean_ndvi, 10)) %>%
  summarise(
    mean_lst_top10 = mean(mean_lst[ndvi_qtile == 10]),
    mean_lst_bottom10 = mean(mean_lst[ndvi_qtile == 1]),
    lst_diff = mean_lst_bottom10 - mean_lst_top10,
    .groups = "drop"
  ) %>%
  arrange(desc(lst_diff))

knitr::kable(quartile_summary, digits = 2, caption = "Temperature Difference: Bottom vs. Top 10% NDVI Tracts")
```

# Census Analysis

*NOTE* 
*To load the needed data, I've used this API key from the US Census Bureau: "census_api_key("ac082ac1a88527ca0092581d65b3a01ae9967027", install = TRUE, overwrite = TRUE)"*

# *8. Dowloading and loading census data*
```{r cens_load, warning=FALSE, message=FALSE, results='hide'}

states_needed <- c("NY", "CA", "IL", "TX", "AZ", "PA")

income_data <- get_acs(geography = "tract",
                       variables = "B19013_001",
                       state = states_needed,
                       year = 2020,
                       geometry = FALSE,
                       output = "wide")
income_data <- income_data %>%
  dplyr::select(GEOID, median_income = B19013_001E)

zonal_stats <- read.csv("data/ndvi_lst_by_tract.csv")
zonal_stats$GEOID <- as.character(zonal_stats$GEOID)
zonal_income <- zonal_stats %>%
  left_join(income_data, by = "GEOID")
```

# *9. Decile Comparison*
```{r cens_dec_graph ,message=FALSE, warning=FALSE}
quartiles <- zonal_income %>%
  filter(!is.na(mean_ndvi), !is.na(mean_lst)) %>%
  group_by(city) %>%
  mutate(ndvi_decile = ntile(mean_ndvi, 10)) %>%
  group_by(city, ndvi_decile) %>%
  summarise(mean_lst = mean(mean_lst), .groups = "drop")

ggplot(quartiles, aes(x = ndvi_decile, y = mean_lst)) +
  geom_line(size = 1.2, color = "#3a8d63") + 
  geom_point(size = 2, color = "#3a8d63") +
  facet_wrap(~ city, scales = "free_y") +  # One plot per city
  labs(title = "LST across NDVI Deciles per City",
       subtitle = "How urban heat varies by green space",
       x = "NDVI Decile (1 = Least Green, 10 = Most Green)",
       y = "Mean LST (°C)") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(size = 14),  # Larger city names
        axis.text = element_text(size = 10))
```

# *10. Income vs NDVI by city*
```{r cens_corr, message=FALSE, warning=FALSE}
ggplot(zonal_income %>% filter(!is.na(mean_ndvi), !is.na(median_income)),
       aes(x = median_income, y = mean_ndvi)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  facet_wrap(~ city, scales = "free_x") + 
  labs(title = "Income vs NDVI by City",
       x = "Median Household Income ($)",
       y = "Mean NDVI") +
  theme_minimal(base_size = 12)
```


**Note:** Median household income data were available for a subset of cities and tracts. The Income–NDVI analysis is therefore shown only for those cities with sufficient coverage.










